---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pandoc

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN
status](https://www.r-pkg.org/badges/version/pandoc)](https://CRAN.R-project.org/package=pandoc)
[![R-CMD-check](https://github.com/cderv/pandoc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/cderv/pandoc/actions/workflows/R-CMD-check.yaml)
[![Codecov test
coverage](https://codecov.io/gh/cderv/pandoc/branch/main/graph/badge.svg)](https://app.codecov.io/gh/cderv/pandoc?branch=main)

<!-- badges: end -->

```{r child = "vignettes/articles/_about.Rmd"}
```

## Usage

All functions are prefixed with `pandoc_` and you will find below their
main usage. See [Get started](articles/pandoc.html) for usage examples
and [Reference](reference/index.html) page for full details of
functions.

Most functions follows these following rules:

-   They are prefixed with `pandoc_` .
-   They can be used with any installed Pandoc version using `version`
    argument.
-   Versions to use are passed as a string, either the version number
    (e.g `"2.14.2"` ), or one special alias (`default`, `nightly` ,
    `rstudio` , `system`) .

Available functions allows:

-   Installing / Uninstalling specific Pandoc version from
    `r knitr::combine_words(range(as.numeric_version(pandoc::pandoc_available_releases())), sep = ", ", and = " to ", before = "'")`
    , including development version of Pandoc.
    -   `pandoc_install()`, `pandoc_install_nightly()`,
        `pandoc_uninstall()`, ...
-   Switching Pandoc version by activating a specific one or running any
    function with a specific Pandoc version using `version=` argument.
    -   `pandoc_activate()` , `with_pandoc()` , ....
-   Managing locally installed version of Pandoc as versions installed
    by this package are located within one folder per version in a
    user’s data directory.
    -   `pandoc_installed_versions()` , `pandoc_installed_latest` ,
        `pandoc::pandoc_is_installed()` , `pandoc_locate()`,
        `pandoc_available()`, ...
-   Using easily one of the installed version with **rmarkdown** (i.e
    `rmarkdown::render()` will use the version activated by this pandoc)
    -   See `pandoc::pandoc_activate(rmarkdown = TRUE)`
-   Running a Pandoc binary from R, including a version installed
    system-wise (`pandoc_bin("system")`), or the version shipped with
    RStudio (`pandoc_bin("rstudio")`).
-   Calling binary at low-level from R with any version installed.
    -   `pandoc_run()`
-   to easily access information in R usually requiring command line
    execution
    -   `pandoc_version()`, `pandoc_list_extensions()` ,
        `pandoc_export_template()` , and all other wrappers...
-   opening Pandoc's resources from R
    -   `pandoc::pandoc_browse_manual()`,
        `pandoc::pandoc_browse_extension()` , ....

## Example

```{r}
library(pandoc)
# Install version
pandoc_install("2.7.3")
pandoc_install("2.11.4")
# Highest install is used
pandoc_version()
```

See detailed examples in [Get started](articles/pandoc.html).

## Code of Conduct

Please note that the **pandoc** project is released with a [Contributor
Code of Conduct](https://cderv.github.io/pandoc/CODE_OF_CONDUCT.html).
By contributing to this project, you agree to abide by its terms.
