---
title: "Comparing Pandoc versions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_template$set(quiet.rlang = list(
  R.options = list(rlib_message_verbosity = "quiet")
))
```

## Why compare Pandoc versions ? 

One of the aim for this package is to help dealing with Pandoc changes over time. Pandoc is a project which has often some breaking change between versions. However, tools like R Markdown aims to work with any Pandoc versions, and hide those changes for the user by insure backward compatibility and adjustement toward new features. 

Dealing with this can require to compare versions and now when a change occurs. 

## Use Case 1: Finding which version introduced a change

Let's take the example of a change in Pandoc regaring `gfm` format: It appears at some point that raw HTML `<span></span>` was removed during conversion. When does this happened ? Let's take a look. 

For that, we want to convert `"This is a <span>Span</span>."` to `gfm` format and look for the max version that keep the raw HTML `<span>` in its markdown output.

```{r setup}
library(pandoc)
```

As a prerequisite to the analysis, all available pandoc versions needs to be installed.

```{r, opts.label = "quiet.rlang"}
purrr::walk(pandoc_available_releases(), purrr::safely(pandoc_install))
```

Once this is done, it is easy to iterate over versions to do an analysis over an hypothesis.

First, we use each Pandoc version to convert `"This is a <span>Span</span>."` to Github Markdown

```{r}
# Get the available versions
versions <- pandoc_installed_versions()
versions <- purrr::set_names(versions)

# Do conversion for each version
res <- purrr::map(
  versions, 
  ~ pandoc_convert(text = "This is a <span>Span</span>.", to = "gfm", version = .x)
)
```

With the result, we can build a tibble to filter out which is the maximum version that preserve the raw `<span>`

```{r}
library(dplyr)
tab <- res %>%
  purrr::map_chr(as.character) %>%
  tibble::enframe("ver", "string")

tab %>%
  # No nightly
  filter(ver != "nightly") %>%
  # converting to numeric version for easier ordering
  mutate(num_ver = as.numeric_version(ver)) %>%
  # Which version does keep <span> ?
  filter(grepl("<span>", string, fixed = TRUE)) %>%
  # Order version
  arrange(num_ver) %>%
  # which is the max one ? 
  slice_tail(n = 1)
```
We can confirm looking at 2.10 and 2.10.1, that `<span>` is indeed not preserve.

```{r}
tab %>%
  filter(ver %in% c("2.10", "2.10.1"))
```
From there, it is easier to know how to adapt, by first looking into the changes in Pandoc's release not 

```{r}
pandoc_browse_release("2.10.1")
```



