---
title: "pandoc"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# run isolated as in test
Sys.setenv(TESTTHAT = "true")

# default pandoc use with rmarkdown
pandoc_default <- rmarkdown::find_pandoc()$dir
```

```{r setup}
library(pandoc)
```

## About this package

This is currently an experimental package aiming mainly to help maintainers of R Markdown ecosystem. 

The R Markdown ecosystem is highly dependent on Pandoc <https://pandoc.org/> and it is designed to be as version independent as possible. R Markdown is best used with the latest Pandoc version but any **rmarkdown** package version should work with previous version, and new change in Pandoc should not break any rmarkdown features.

This explain the needs to some more focused tooling to:

* Install and manage several pandoc versions. This is useful for testing versions and comparing between them.
* Call pandoc directly without the layers added by **rmarkdown**. This is useful for debugging or quickly iterating and finding where a bug could be. 
* Retrieve information from Pandoc directly. Each version comes with changes and some of them are included into the binary. Being able to retrieve those information and compare between version is also important to help maintain the user exposed tooling.

This can also be useful to advanced developer that are working around pandoc through **rmarkdown** or not. 

## Install pandoc

Main usage is to install latest version of Pandoc. This requires the `gh` package as it will fetch information from Github and download the bundle from there.

```{r}
pandoc_install()
```

If a specific older Pandoc version is needed (e.g for testing differences between version), a version can be specified. 
```{r}
pandoc_install("2.11.4")
```

Informations fetched from Github are cached for the session.

Sometimes, the dev version of Pandoc is required. Pandoc's team is building a binary every day called _nightly_
```{r}
pandoc_install_nightly() # or pandoc_install("nightly")
```

All those version can live together and are installed in an isolated directory in user's data folders.

```{r}
# Which version are currently installed ? 
pandoc_installed_versions()
# Which is the latest version installed (nightly excluded)? 
pandoc_installed_latest()
# Is a specific version installed ? 
pandoc_is_installed("2.11.4")
pandoc_is_installed("2.7.3")
```

Downloaded bundles are also cached to speed up further installation. This is useful into tests to quickly install and uninstall a pandoc version for a specific test. 

```{r}
pandoc_install("2.7.3")
pandoc_uninstall("2.7.3")
pandoc_install("2.7.3")
```

To quickly install the last available release, run `pandoc_update()` (or equivalent `pandoc_install()`).

## Find where a pandoc binary is located

For any version installed with this package, `pandoc_locate()` will return the folder where it was installed. 

```{r}
pandoc_locate("2.11.4")
pandoc_locate("nightly")
```

For examples purposes in this vignette, the path above is in a temp directory. Correct location is in user's data directory computed with `rappdirs::user_data_dir()` (e.g on Windows `C:/Users/chris/AppData/Local/r-pandoc/r-pandoc`)

To the path to a pandoc binary, `pandoc_bin()` can be used
```{r}
pandoc_bin("2.11.4")
pandoc_bin("nightly")
```

This function also bring support for external pandoc version.  like 

* the one shipped with RStudio IDE (`version = "rstudio")`: 
  * `pandoc::pandoc_bin("rstudio")`
  * `pandoc::pandoc_rstudio_bin()` 
* the one set by default on the system (in PATH) (`version = "system"`):
  * `pandoc::pandoc_bin("system")`
  * `pandoc::pandoc_system_bin()`
  
## Activate a Pandoc version

As multiple versions can be installed, a default active pandoc version will be used with any of the function. (`version = "default"`). 

A specific version can be made active using `pandoc_activate()`

```{r}
# Default to latest version installed
pandoc_activate()
pandoc_locate()
pandoc_bin()

# Activate specific version
pandoc_activate("2.11.4")
pandoc_locate()
pandoc_bin()

# including nightly
pandoc_activate("nightly")
pandoc_locate()
pandoc_bin()

# Activate system version
pandoc_activate("system")
pandoc_bin()
```

A default active version will be set when the package is loaded (i.e using `onLoad`) following this search order: 

* Latest version install by this package (i.e `pandoc_installed_latest()`)
* Version shipped with RStudio IDE (found when run inside RStudio IDE)
* pandoc binary found in system PATH  (i.e `Sys.which("pandoc")`)

`pandoc_is_active()` allows to easily know if a specific version is active or not.

```{r}
pandoc_is_active("system")
pandoc_is_active("2.7.3")
pandoc_activate("2.7.3")
pandoc_is_active("2.7.3")
```

### Working with **rmarkdown** functions related to Pandoc

By default, if **rmarkdown** is installed, `pandoc_activate()` will also set the version active for all **rmarkdown** functions (using `rmarkdown::find_pandoc()`). This allows to use this package easily in order to test **rmarkdown** with different version of Pandoc.

```{r}
pandoc_activate("2.7.3")
rmarkdown::pandoc_available()
rmarkdown::pandoc_version()
rmarkdown::find_pandoc()
```

These calls are equivalent
```{r}
pandoc_activate("2.7.3", rmarkdown = TRUE)
rmarkdown::find_pandoc(cache = FALSE, dir = pandoc::pandoc_locate("2.7.3"))
```

If setting the default Pandoc version for **rmarkdown** is not desired, just run with `rmarkdown = FALSE`
```{r}
pandoc::pandoc_activate("2.11.4", rmarkdown = FALSE)
rmarkdown::pandoc_version()
```

During testing, it also interesting to run a specific code with a specific version. `with_pandoc_version()` or `local_pandoc_version()` allows by running `pandoc_activate()` for the expression only (helper like [withr](https://withr.r-lib.org/)). 

```{r}
rmarkdown::pandoc_version()

with_pandoc_version("2.11.4", {
  rmarkdown::pandoc_version()
})
with_pandoc_version("2.7.3", {
  rmarkdown::pandoc_version()
})
```

```{r, include=FALSE}
# restore rmarkdown's pandoc
rmarkdown::find_pandoc(cache = FALSE, dir = pandoc_default)
```

### Check if a pandoc version is available

```{r}
# Is a pandoc version available (i.e a version is active), and if so what is the full path ? 
if (pandoc_available()) pandoc_bin()

# Is the active version above 2.10.1 ?
pandoc_available(min = "2.10.1")
# Is the active version below 2.11 ?
pandoc_available(max = "2.11")
# Is the active version between 2.10.1 and 2.11, both side include ?
pandoc_available(min = "2.10.1", max = "2.11")
```

Pandoc version can also easily be retrieved, including for external binaries

```{r}
# Get version from current active one
pandoc_version()
# Get version for a specific version
pandoc_version("nightly")
# Get version for a specific version
pandoc_version("system") # equivalent to pandoc_system_version()
```

## Run Pandoc CLI

### Low level call to Pandoc

`pandoc_run()` is the function to call pandoc binary with some arguments. By default, it will use the active version (`version = "default"`, see `?pandoc_activate`)

```{r}
pandoc_run("--version")
```
equivalent to calling
```bash
pandoc --version
```
with the correct binary.

Using the `version=` argument allows to run a specific version
```r
pandoc_run("--version", version = "system")
```
will execute the pandoc command with pandoc binary on PATH.

### Convert a document

> This function is highly experimentatl and probability of API change is high.

Main usage of Pandoc is to convert a document. The `pandoc::pandoc_convert()` is currently a thinner wrapper than `rmarkdown::pandoc_convert()`. Both allow to convert a file but the former also allow to convert from text

```{r}
# convert from text directly
pandoc_convert(text = "# A header", to = "html")
pandoc_convert(text = "# A header", to = "html", version = "system") 

# convert from file
tmp <- tempfile(fileext = ".md")
writeLines("**bold** word!", tmp)
pandoc_convert(tmp, to = "html")
# write to file
out <- tempfile(fileext = ".html")
outfile <- pandoc_convert(tmp, to = "html", output = out, standalone = TRUE, version = "system")
readLines(outfile, n = 5)
```

### Various Wrapper functions around pandoc CLI

All other included functions to run pandoc are wrapping `pandoc_run()` with some command flags form [Pandoc MANUAL](https://pandoc.org/MANUAL.html).
Each of these function can take the `version=` argument to run with a specific version of Pandoc instead of the current activated one.

Some of those functions can only be used with specific pandoc versions and an error will be thrown if the version requirement is not met.

#### List supported extensions for a format

```{r}
pandoc_list_extensions()
pandoc_list_extensions(format = "gfm")
pandoc_list_extensions(format = "html", version = "nightly")
```

#### List available input or output formats

```{r}
pandoc_list_formats("input")
pandoc_list_formats("output")
pandoc_list_formats("output", version = "nightly")
```

#### List available highlight style

```{r}
pandoc_list_highlight_style()
```

#### List supported highlight language

```{r}
pandoc_list_highlight_languages()
```

#### Export a data file

```{r}
outfile <- pandoc_export_data_file(file = "styles.html")
outfile
readLines(outfile, n = 5)
```

```{r, include = FALSE}
unlink(outfile)
```

#### Export a highlight style JSON file

```{r}
outfile <- pandoc_export_highlight_theme(style = "zenburn")
outfile
readLines(outfile, n = 5)
```

```{r, include = FALSE}
unlink(outfile)
```

#### Export a DOCX or PTTX reference doc

```{r}
ref_docx <- pandoc_export_reference_doc(type = "docx")
ref_docx
ref_pptx <- pandoc_export_reference_doc(type = "pptx")
ref_pptx
```

```{r, include = FALSE}
unlink(c(ref_docx, ref_pptx))
```

#### Export a template for a format

```{r}
pandoc_export_template(format = "jira")
outfile <- pandoc_export_template(format = "latex",output = "default.latex")
outfile
readLines(outfile, n = 5)
```

```{r, include = FALSE}
unlink(c(outfile))
```

## Helpers to easily browse Pandoc's online resources 

`pandoc_browse_*()` helpers are included to quickly open an online document like the Pandoc MANUAL (`pandoc_browse_manual()`) or a documentation for an extensions (`pandoc_browse_extension("smart")`). See [reference doc](reference/dev/reference/index.html#section-browse-pandoc-s-useful-online-resources) for more.
